The Talking Heads Experiment

### 2015-04-27 chap10/chap10.tex
attended to all todonotes

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap10/figs/look-ball.pdf}
\includegraphics[width=.45\textwidth]{chap10/figs/views-ball.pdf}}
\caption{\label{f:look-ball}Left: Fr\'{e}d\'{e}ric Kaplan interacts with the \textsc{aibo} robot in an experiment for the social learning of lexicons.
Right: View of the ball through the eye of the robot. As the same object is now seen from many different viewpoints it 
becomes much more challenging to consistently recognise it.}
\end{figure}
The first \textsc{aibo} experiments did not involve a population of agents but rather a single agent that was interacting directly with a human
experimenter. The agent had to learn the lexicon of the human through social 
interaction within the context of situated language games. The 2001
paper ``\textsc{aibo}'s first words'' by Steels and Kaplan \todo{citation?} described experiments in which an \textsc{aibo} robot 
learned words about objects in the environment, such as `ball' or `smiley', as well as words for actions, such as 
`sit down' or `stand up'.

@@

``The grammar is seen as a natural continuation of the lexicon, in the sense that it consists also of associations between forms and meanings. Use and success are monitored for each association so that the same type of self-organised coherence arises in the group, as seen in the lexicon. The form is now a more complex structure, defined as a syntactic schema. The meaning is a semantic schema. The schemas
circumscribe a feature-set in terms of a set of slots, restrictions	
on the fillers of each slot, and constraints on the combination	
of the fillers to form the total covered by the schema. \todo{This quote does not have an end}

Syntactic schemas describe word groups. They have an associated category which corresponds in linguistic terms to grouping categories like noun-group, verb-group, sentence. The slots in syntactic schemas correspond to syntactic functions (also called grammatical relations) such as subject, object, modifier, complement. They name the roles that certain words or word groups play in the group. The categories used to restrict possible slot-fillers correspond in linguistic terms to syntactic categories like noun, verb, adjective, etc. An example of a syntactic schema is the following:
\begin{verbatim}
Schema-541 
  SLOTS: (syn-slot-51 syn-slot-50) 
  DESCRIPTION-SET: 
      ([syn-slot-50 syn-cat-751] 
       [syn-slot-51 syn-cat-761])
  CONSTRAINTS:
      ((PRECEEDS (>> syn-slot-50) (>> syn-slot-51)))
  CATEGORY: syn-cat-77 
  USE: 10 
  SUCCESS: 3
\end{verbatim}
Of course, instead of noun, noun-group, subject, etc. computer-generated symbols are used such 
as syn-cat-77, syn-slot-50, etc. The constraints on the schema are represented in terms of
a constraint system. Each constraint
has a dual procedural function: to enforce the constraint when re-enacting the situation
described by a schema or to test the constraint when determining whether a schema fits with a situation. In the present
case only a precedence relation is recorded as a constraint. Agreement, intonation, or morphological
endings, are some other possible constraints on syntactic schemas.

>>>>

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap10/figs/look-ball.pdf}
\includegraphics[width=.45\textwidth]{chap10/figs/views-ball.pdf}}
\caption{\label{f:look-ball}Left: Fr\'{e}d\'{e}ric Kaplan interacts with the \textsc{aibo} robot in an experiment for the social learning of lexicons.
Right: View of the ball through the eye of the robot. As the same object is now seen from many different viewpoints it 
becomes much more challenging to consistently recognise it.}
\end{figure}
The first \textsc{aibo} experiments did not involve a population of agents but rather a single agent that was interacting directly with a human
experimenter. The agent had to learn the lexicon of the human through social 
interaction within the context of situated language games. The 2001
paper ``\textsc{aibo}'s first words'' by Steels and Kaplan \nocite{Steels:01h} described experiments in which an \textsc{aibo} robot 
learned words about objects in the environment, such as `ball' or `smiley', as well as words for actions, such as 
`sit down' or `stand up'.

@@

\begin{quote}
The grammar is seen as a natural continuation of the lexicon, in the sense that it consists also of associations between forms and meanings. Use and success are monitored for each association so that the same type of self-organised coherence arises in the group, as seen in the lexicon. The form is now a more complex structure, defined as a syntactic schema. The meaning is a semantic schema. The schemas
circumscribe a feature-set in terms of a set of slots, restrictions	
on the fillers of each slot, and constraints on the combination	
of the fillers to form the total covered by the schema.

Syntactic schemas describe word groups. They have an associated category which corresponds in linguistic terms to grouping categories like noun-group, verb-group, sentence. The slots in syntactic schemas correspond to syntactic functions (also called grammatical relations) such as subject, object, modifier, complement. They name the roles that certain words or word groups play in the group. The categories used to restrict possible slot-fillers correspond in linguistic terms to syntactic categories like noun, verb, adjective, etc. An example of a syntactic schema is the following:
\begin{verbatim}
Schema-541 
  SLOTS: (syn-slot-51 syn-slot-50) 
  DESCRIPTION-SET: 
      ([syn-slot-50 syn-cat-751] 
       [syn-slot-51 syn-cat-761])
  CONSTRAINTS:
      ((PRECEEDS (>> syn-slot-50) (>> syn-slot-51)))
  CATEGORY: syn-cat-77 
  USE: 10 
  SUCCESS: 3
\end{verbatim}
\end{quote}

Of course, instead of noun, noun-group, subject, etc. computer-generated symbols are used such 
as syn-cat-77, syn-slot-50, etc. The constraints on the schema are represented in terms of
a constraint system. Each constraint
has a dual procedural function: to enforce the constraint when re-enacting the situation
described by a schema or to test the constraint when determining whether a schema fits with a situation. In the present
case only a precedence relation is recorded as a constraint. Agreement, intonation, or morphological
endings, are some other possible constraints on syntactic schemas.

### 2015-04-27 chap11/chap11.tex
attended to all todonotes

\subsection{Semiotic networks}\todo{This used to be set in a \textbackslash bfshape and not in a subsection}
We have already seen from earlier experiments that the solution to these various issues lies in setting up
the right kind of ``semiotic dynamics''. In the present case, this dynamics should gradually coordinate
sensory experiences, prototypical views, individuals, 
and names, both within a single individual and across the
population. This lead to the notion of a {\itshape semiotic network}, which was a significant conceptual advance 
in the methodology of conceiving and carrying out language game experiments.

>>>>

\subsection{Semiotic networks}
We have already seen from earlier experiments that the solution to these various issues lies in setting up
the right kind of ``semiotic dynamics''. In the present case, this dynamics should gradually coordinate
sensory experiences, prototypical views, individuals, 
and names, both within a single individual and across the
population. This lead to the notion of a {\itshape semiotic network}, which was a significant conceptual advance 
in the methodology of conceiving and carrying out language game experiments.

### 2015-04-27 chap12/chap12.tex
attended to all todonotes

Today, Darwinian evolutionary dynamics \is{Darwinian evolutionary dynamics}
is recognised as a general principle that can and has been applied to 
many different types of systems, in chemistry, neurobiology or 
economics.\footnote{See \cite{Arthur:1996}, \cite{Edelman:1987} and \cite{Luisi:2001}.}
It has also been used successfully to come up 
with new artificial systems, for example to evolve controllers for robots using 
genetic algorithms \citep{Nolfi:2000}. It is therefore not so surprising that 
selectionism can also be mapped to the cultural and more specifically to the linguistic domain in the following way: 
\begin{enumerate}
\item Speakers generate variants, either accidentally through errors in their speech production, or deliberately, 
for example, when they try to express a novel idea by inventing new forms or 
appropriating existing forms for new purposes, for example, by coercing a word or grammatical contruction into a new usage. 
\item Which variants will survive in the communal language depends on selectionist criteria. I will 
discuss them shortly in more detail. They include for example whether the speaker is understood by the hearer, or whether 
it reduces the cognitive complexity of parsing and producing and hence allows larger sentences to be understood. 
\item The self-enforcing causal loop happens when variants created by one speaker start to propagate in the population 
and begin to replace their competitors. For example, a particular word to express a given meaning gets out of fashion 
and is taken over by another one. Some of these developments are simply fashion trends, but the more profound ones, 
in particular the ones that involve grammar, are related to critical selectionist criteria. 
\end{enumerate}
This Darwinian dynamic is already active in the operation of language strategies, since \todo{replaced `because' with `since' here to fit the line} a strategy regulates the 
creation and selection of variants, determines how feedback has to be handled, and implies a self-enforcing causal loop through 
lateral inhibition so that variants satisfying the selection criteria become 
dominant in the population.

@@

Recall that three internal factors were used \todo{replaced `started' with `began' here to fit the line}
for the fitness landscape in this domain: communicative success, semantic ambiguity and interpretation sharing. Spranger began with 
a population without any grammar but just a lexicon for expressing spatial relations and properties of objects. 
The agents were then initialised with a strategy that would build a grammar consisting of hierarchical
constructions which group all the components of a spatial relation, put them in a particular sequential order,
and add a marker to them. For example, a construction could arise that groups a spatial relation (like `front'), 
a determiner acting as selector (like `the'), and a noun introducing an object class (e.g. `box'), puts them 
in the order: spatial relation - determiner - noun, and adds a marker to them (e.g. `-bo').

>>>>

Today, Darwinian evolutionary dynamics \is{Darwinian evolutionary dynamics}
is recognised as a general principle that can and has been applied to 
many different types of systems, in chemistry, neurobiology or 
economics.\footnote{See \cite{Arthur:1996}, \cite{Edelman:1987} and \cite{Luisi:2001}.}
It has also been used successfully to come up 
with new artificial systems, for example to evolve controllers for robots using 
genetic algorithms \citep{Nolfi:2000}. It is therefore not so surprising that 
selectionism can also be mapped to the cultural and more specifically to the linguistic domain in the following way: 
\begin{enumerate}
\item Speakers generate variants, either accidentally through errors in their speech production, or deliberately, 
for example, when they try to express a novel idea by inventing new forms or 
appropriating existing forms for new purposes, for example, by coercing a word or grammatical contruction into a new usage. 
\item Which variants will survive in the communal language depends on selectionist criteria. I will 
discuss them shortly in more detail. They include for example whether the speaker is understood by the hearer, or whether 
it reduces the cognitive complexity of parsing and producing and hence allows larger sentences to be understood. 
\item The self-enforcing causal loop happens when variants created by one speaker start to propagate in the population 
and begin to replace their competitors. For example, a particular word to express a given meaning gets out of fashion 
and is taken over by another one. Some of these developments are simply fashion trends, but the more profound ones, 
in particular the ones that involve grammar, are related to critical selectionist criteria. 
\end{enumerate}
This Darwinian dynamic is already active in the operation of language strategies, since  a strategy regulates the 
creation and selection of variants, determines how feedback has to be handled, and implies a self-enforcing causal loop through 
lateral inhibition so that variants satisfying the selection criteria become 
dominant in the population.

@@

Recall that three internal factors were used 
for the fitness landscape in this domain: communicative success, semantic ambiguity and interpretation sharing. Spranger began with 
a population without any grammar but just a lexicon for expressing spatial relations and properties of objects. 
The agents were then initialised with a strategy that would build a grammar consisting of hierarchical
constructions which group all the components of a spatial relation, put them in a particular sequential order,
and add a marker to them. For example, a construction could arise that groups a spatial relation (like `front'), 
a determiner acting as selector (like `the'), and a noun introducing an object class (e.g. `box'), puts them 
in the order: spatial relation - determiner - noun, and adds a marker to them (e.g. `-bo').

### 2015-04-27 chap2/chap2.tex
attended to all todonotes

The teleporting setup enables \todo{'some' was deleted here to fit the line} fascinating 
experiments, engulfing the 
whole globe. The same agent can look at a scene from
different points of view or at scenes in different
physical locations in the world by teleporting himself in different 
bodies. He can develop categories in one location
and enrich his learning experience by moving to a new 
location which has different objects and thus poses new 
categorial challenges.
There can be populations of varying sizes with new 
agents being born and older agents dying, just like in 
natural human populations, so that we can study the 
transmission of language from one generation to the next
or the resilience of a language against an influx and 
outflux of agents. We can also let agents develop in
different parts of the world and have them migrate to
study intercultural exchange and language contact.

@@

Agents can only engage in interactions with other agents 
when they are physically instantiated
in a robot. Each agent has a basic brain architecture
with different layers performing the cognitive functions 
relevant for playing language games: 
\begin{itemize}
\item A perceptual layer which performs low-level signal 
processing to segment the image and collect data about 
each segment such as the colour, size, position or shape
of a segment. 
\item A conceptual layer which categorises and 
conceptualises the segmented and processed image. 
It is based on a self-generated and evolving repertoire
of categorial distinctions, such as red versus green, or
small versus large. Such a repertoire is referred to as 
the agent's ontology in this book. 
\item A lexical layer which maintains an evolving repertoire
of associations between meanings and words, which I will refer to 
as the lexicon, and performs lexical lookup while parsing 
or producing utterances. 
\item A syntactic layer which uses 
grammatical schemata for organising words
in larger structures or for recognising these structures and
reconstructing complex meanings. 
\item A pragmatic layer which carries out the
scripts for playing language games and maintains the machinery 
for engaging in interactions with other agents in a 
shared environment. 
\end{itemize}
Each of these layers is described in more detail 
later. The internals of the layers are not static but
constantly evolving and adapting. They are not strictly 
modular but coupled in various ways to each other.
Each verbal interaction 
in effect changes the agent's internal state and thus 
influences future behaviour. A new, virgin agent 
starts without any built-in ontology, lexicon, or grammar. 
This is one of the crucial points of the whole experiment, 
because we want to test possible theories on how language
and meaning can evolve and be acquired ab initio\todo{removed italics here}.

@@

Each of these activities requires knowledge and/or skill
(summarised in the table below). Perceiving requires 
visual processes capable of segmenting images and deriving image 
segments. Conceptualisation requires an ontology, 
a repertoire of perceptually grounded distinctions
that can be applied to a segmented image to yield 
distinctive categories or category combinations that 
may constitute the meaning of the utterance. Verbalising 
a conceptualisation requires 
a {\itshape lexicon} that maps parts of the meaning to words and 
a {\itshape syntax} that specifies how to organise individual words
into a larger complex. 
The hearer must use similar 
knowledge sources in the other direction. He must use 
his lexicon and syntax to reconstruct the meanings expressed
by the utterance, and then use the 
ontology again to apply the meaning to the present 
context to find the referent (see \tabref{tab:2:findthereferent}).\todo{I inserted `(see ...)' here)}

\begin{table}[htbp]
\caption{}
\label{tab:2:findthereferent}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Activity} & {\itshape From} & {\itshape To} & {\itshape Knowledge source}\\ \midrule
Perceive & Object & Image segments & Visual processes \\ 
Conceptualise & Topic & Meaning & Ontology  \\ 
Verbalise & Meaning & Utterance & Lexicon + syntax \\  
Interpret & Utterance & Meaning & Lexicon + syntax \\ 
Apply & Meaning & Topic & Ontology \\ 
Act & Topic & Object & Behavioural process  \\ 
\lspbottomrule
\end{tabular}
\end{table}
\todo{Table 2.1 table needs a caption}
There is no simple linear flow from perception to utterance 
or from utterance to perception. Instead, we must imagine a dynamic process 
involving forward and backward propagation of information 
until coherent choices for all the nodes in the two
semiotic squares have been established by the speaker and 
the hearer. Many different choices are initially possible 
(many segmentations, conceptualisations, verbalisations, 
interpretations), but the dynamic process gradually settles into 
a single coherent attractor, so that speaker and hearer 
agree upon a common referent. The co-ordination between 
chosen topic and identified referent is done through 
the real world (pointing, handing an object, performing
an action).

@@

The {\itshape semantic theory} required to make the 
Talking Heads experiment work is very different from 
the classical textbook approaches to the subject, which
tend to assume that categories can be defined in the 
abstract, independent of their context of use. Grounded
processes of conceptualisation and interpretation are necessarily \todo{Word order changed to fit line.}
strongly situated and context-dependent. Whether `blue 
triangle' is going to be effective in picking out the 
intended topic depends on what else is in
the scene. If all other objects in the 
context are also blue triangles, the game will fail. If there
is only one triangle, it was not really necessary to say that 
it is blue. This situatedness and context-dependence, together with 
the non-verbal hints given by the speaker, are crucial 
keys for restricting the 
possible meanings of a form 
and they therefore help the hearer disambiguate 
utterances or figure out the meaning of an unknown form. But situatedness 
and con\-text-dependence are also major
sources of difficulty, because the same categories may
refer to different things for different agents in different
contexts. Even if both agents agree that a particular form 
names a certain category, 
they may still fail in the game if the category picks
out a different referent, for example because they see 
the context from a slightly different vantage point.

>>>>

The teleporting setup enables fascinating 
experiments, engulfing the 
whole globe. The same agent can look at a scene from
different points of view or at scenes in different
physical locations in the world by teleporting himself in different 
bodies. He can develop categories in one location
and enrich his learning experience by moving to a new 
location which has different objects and thus poses new 
categorial challenges.
There can be populations of varying sizes with new 
agents being born and older agents dying, just like in 
natural human populations, so that we can study the 
transmission of language from one generation to the next
or the resilience of a language against an influx and 
outflux of agents. We can also let agents develop in
different parts of the world and have them migrate to
study intercultural exchange and language contact.

@@

Agents can only engage in interactions with other agents 
when they are physically instantiated
in a robot. Each agent has a basic brain architecture
with different layers performing the cognitive functions 
relevant for playing language games: 
\begin{itemize}
\item A perceptual layer which performs low-level signal 
processing to segment the image and collect data about 
each segment such as the colour, size, position or shape
of a segment. 
\item A conceptual layer which categorises and 
conceptualises the segmented and processed image. 
It is based on a self-generated and evolving repertoire
of categorial distinctions, such as red versus green, or
small versus large. Such a repertoire is referred to as 
the agent's ontology in this book. 
\item A lexical layer which maintains an evolving repertoire
of associations between meanings and words, which I will refer to 
as the lexicon, and performs lexical lookup while parsing 
or producing utterances. 
\item A syntactic layer which uses 
grammatical schemata for organising words
in larger structures or for recognising these structures and
reconstructing complex meanings. 
\item A pragmatic layer which carries out the
scripts for playing language games and maintains the machinery 
for engaging in interactions with other agents in a 
shared environment. 
\end{itemize}
Each of these layers is described in more detail 
later. The internals of the layers are not static but
constantly evolving and adapting. They are not strictly 
modular but coupled in various ways to each other.
Each verbal interaction 
in effect changes the agent's internal state and thus 
influences future behaviour. A new, virgin agent 
starts without any built-in ontology, lexicon, or grammar. 
This is one of the crucial points of the whole experiment, 
because we want to test possible theories on how language
and meaning can evolve and be acquired {\itshape ab initio}.

@@

Each of these activities requires knowledge and/or skill
(summarised in the table below). Perceiving requires 
visual processes capable of segmenting images and deriving image 
segments. Conceptualisation requires an ontology, 
a repertoire of perceptually grounded distinctions
that can be applied to a segmented image to yield 
distinctive categories or category combinations that 
may constitute the meaning of the utterance. Verbalising 
a conceptualisation requires 
a {\itshape lexicon} that maps parts of the meaning to words and 
a {\itshape syntax} that specifies how to organise individual words
into a larger complex. 
The hearer must use similar 
knowledge sources in the other direction. He must use 
his lexicon and syntax to reconstruct the meanings expressed
by the utterance, and then use the 
ontology again to apply the meaning to the present 
context to find the referent (see \tabref{tab:2:findthereferent}).

\begin{table}[htbp]
\caption{Activities and knowledge soruces}
\label{tab:2:findthereferent}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Activity} & {\itshape From} & {\itshape To} & {\itshape Knowledge source}\\ \midrule
Perceive & Object & Image segments & Visual processes \\ 
Conceptualise & Topic & Meaning & Ontology  \\ 
Verbalise & Meaning & Utterance & Lexicon + syntax \\  
Interpret & Utterance & Meaning & Lexicon + syntax \\ 
Apply & Meaning & Topic & Ontology \\ 
Act & Topic & Object & Behavioural process  \\ 
\lspbottomrule
\end{tabular}
\end{table}

There is no simple linear flow from perception to utterance 
or from utterance to perception. Instead, we must imagine a dynamic process 
involving forward and backward propagation of information 
until coherent choices for all the nodes in the two
semiotic squares have been established by the speaker and 
the hearer. Many different choices are initially possible 
(many segmentations, conceptualisations, verbalisations, 
interpretations), but the dynamic process gradually settles into 
a single coherent attractor, so that speaker and hearer 
agree upon a common referent. The co-ordination between 
chosen topic and identified referent is done through 
the real world (pointing, handing an object, performing
an action).

@@

The {\itshape semantic theory} required to make the 
Talking Heads experiment work is very different from 
the classical textbook approaches to the subject, which
tend to assume that categories can be defined in the 
abstract, independent of their context of use. Grounded
processes of conceptualisation and interpretation are necessarily
strongly situated and context-dependent. Whether `blue 
triangle' is going to be effective in picking out the 
intended topic depends on what else is in
the scene. If all other objects in the 
context are also blue triangles, the game will fail. If there
is only one triangle, it was not really necessary to say that 
it is blue. This situatedness and context-dependence, together with 
the non-verbal hints given by the speaker, are crucial 
keys for restricting the 
possible meanings of a form 
and they therefore help the hearer disambiguate 
utterances or figure out the meaning of an unknown form. But situatedness 
and con\-text-dependence are also major
sources of difficulty, because the same categories may
refer to different things for different agents in different
contexts. Even if both agents agree that a particular form 
names a certain category, 
they may still fail in the game if the category picks
out a different referent, for example because they see 
the context from a slightly different vantage point.

### 2015-04-27 chap3/chap3.tex
attended to all todonotes

\subsection{Simulation versus experimentation}

Working \todo{3.5.1 is orphaned} with simulations has obvious advantages for 
speeding up development and systematic testing, 
but it does not replace experimentation with 
physical robots. It is true that 
building an experimental apparatus such as we built
for the Talking Heads experiment is a very 
non-trivial engineering project, particularly because 
of the teleporting infrastructure that enables the agents 
to travel to multiple sites and thus 
experience different physical environments or the same 
environment from different points of view. 
One might therefore wonder why we have persisted
to go to such great length in constructing
a real-world physical infrastructure. Why are
experiments with robotic agents necessary or 
desirable? Is it not enough to conduct simulation
experiments given that they can be done so much 
more easily?

>>>>

% \subsection{Simulation versus experimentation}

Working  with simulations has obvious advantages for 
speeding up development and systematic testing, 
but it does not replace experimentation with 
physical robots. It is true that 
building an experimental apparatus such as we built
for the Talking Heads experiment is a very 
non-trivial engineering project, particularly because 
of the teleporting infrastructure that enables the agents 
to travel to multiple sites and thus 
experience different physical environments or the same 
environment from different points of view. 
One might therefore wonder why we have persisted
to go to such great length in constructing
a real-world physical infrastructure. Why are
experiments with robotic agents necessary or 
desirable? Is it not enough to conduct simulation
experiments given that they can be done so much 
more easily?

### 2015-04-27 chap4/chap4.tex
attended to all todonotes

Categorisation has fascinated philosophers and scientists
since the onset \todo{beginning was replaced with onset here to fit the line} of thought, not only because it is one
of the most fundamental capabilities of the human mind, 
but also because the subject matter immediately raises 
some intriguing paradoxes and puzzles. First of all, 
we have already seen that there is an enormous 
gap between the symbolic world of objects and categories
and the subsymbolic world of analog sensori-motor signal streams. 
A particular sensory signal is highly context-dependent and
inherently noisy due to the partial
unreliability of sensors and their limited accuracy. 
Sensory processing, transformation, and scaling 
go some way to achieve perceptual constancy
but cognitive processes must clearly make up for
the fleeting erratic nature of reality. 
Here we are immediately confronted with a first paradox. 
Univocal categorisation seems only possible when 
an interpretation of reality is within reach, for example
when there are strong expectations, possibly coming 
from language utterances. But this interpretation
depends itself on categorisation. How can this apparent
causal circularity be broken?

>>>>

Categorisation has fascinated philosophers and scientists
since the onset of thought, not only because it is one
of the most fundamental capabilities of the human mind, 
but also because the subject matter immediately raises 
some intriguing paradoxes and puzzles. First of all, 
we have already seen that there is an enormous 
gap between the symbolic world of objects and categories
and the subsymbolic world of analog sensori-motor signal streams. 
A particular sensory signal is highly context-dependent and
inherently noisy due to the partial
unreliability of sensors and their limited accuracy. 
Sensory processing, transformation, and scaling 
go some way to achieve perceptual constancy
but cognitive processes must clearly make up for
the fleeting erratic nature of reality. 
Here we are immediately confronted with a first paradox. 
Univocal categorisation seems only possible when 
an interpretation of reality is within reach, for example
when there are strong expectations, possibly coming 
from language utterances. But this interpretation
depends itself on categorisation. How can this apparent
causal circularity be broken?

### 2015-04-27 chap5/chap5.tex
attended to all todonotes

Every time an agent successfully uses a form-meaning pair 
for speaking, he increments
the score with a specific increment $\delta$. 
$\delta$ is relatively high, typically equal to 0.1. 
The scores of competing associations, i.e. 
associations that used another form for the same meaning
are decremented with $\delta$. The score remains however bounded 
between 0.0 and 1.0. This way the ``best'' 
form-meaning pair stands out more clearly next time 
around.\todo{Cite Kaplan's thesis, please}\footnote{A systematic investigation of alternatives for 
the updating function is contained in \cite{Oliphant:1997}. 
The dynamics of the mechanisms used in the 
Talking Heads experiments are being investigated in the
Ph.D thesis of Frederic Kaplan.}

>>>>

Every time an agent successfully uses a form-meaning pair 
for speaking, he increments
the score with a specific increment $\delta$. 
$\delta$ is relatively high, typically equal to 0.1. 
The scores of competing associations, i.e. 
associations that used another form for the same meaning
are decremented with $\delta$. The score remains however bounded 
between 0.0 and 1.0. This way the ``best'' 
form-meaning pair stands out more clearly next time 
around.\footnote{A systematic investigation of alternatives for 
the updating function is contained in \cite{Oliphant:1997}. 
The dynamics of the mechanisms used in the 
Talking Heads experiments are being investigated in the
Ph.D thesis of Frederic Kaplan. \nocite{Kaplan:2001}}

### 2015-04-27 chap6/chap6.tex
attended to all todonotes

We see that words for \todo{`the' deleted here to fit the line} basic distinctions have further established
themselves. Words for `short' (\emph{suwaxugo}) and `tall' (\emph{kusone}), or `light' (\emph{fesasado}) and `dark' (\emph{pokuneso}) now have scores of 1.0. Words for more refined categories, like `very short' (\emph{tawube}) or `very narrow' (\emph{kutaki}), are beginning to establish themselves.

@@

Despite \todo{removed `the' here to fit the line} additional complication of mutually 
compatible meanings for unknown words, the two agents
nevertheless manage to 
build up a shared communication system, as can be seen 
from \figref{gsucc3}. Each time the ontology is 
extended, communicative success dips because a new 
word needs to be acquired, but the agents clearly 
manage to become successful in the guessing game.

>>>>

We see that words for  basic distinctions have further established
themselves. Words for `short' (\emph{suwaxugo}) and `tall' (\emph{kusone}), or `light' (\emph{fesasado}) and `dark' (\emph{pokuneso}) now have scores of 1.0. Words for more refined categories, like `very short' (\emph{tawube}) or `very narrow' (\emph{kutaki}), are beginning to establish themselves.

@@

Despite additional complication of mutually 
compatible meanings for unknown words, the two agents
nevertheless manage to 
build up a shared communication system, as can be seen 
from \figref{gsucc3}. Each time the ontology is 
extended, communicative success dips because a new 
word needs to be acquired, but the agents clearly 
manage to become successful in the guessing game.

### 2015-04-27 chap7/chap7.tex
attended to all todonotes

These competition diagrams are an important tool to try 
and make sense of the ontological and lexical evolution
taking place in evolving groups of agents as they are
playing their language games. Typically we pick
a dominating word, for example the word `va' in 
\figref{RF-diagram}, and try to understand
why it has become dominant. 
The FM-diagram for `va' (\figref{FM-diagram}) explains
part of the story. Three stable meanings for `va' have emerged
at \todo{inserted `at' here} around 1000 games: 
[\textsc{r} 0.0–0.125], [\textsc{b} 0.3125–0.375], and [\textsc{vpos} 0.25–0.5]. 
They are all equally adequate for distinguishing the object
`va' designates, and there are no situations yet
that would have forced the disambiguation of `va'. 
In game 3000, the environment (which is continuously 
changing in this experiment) produces a scene in which
a category which was distinctive for the object designated by 
`va' is no longer distinctive. The lexicon adapts 
immediately. Around game 3000 the \textsc{vpos}-based meaning disappears, 
and the distinction based on \textsc{red} shoots up and becomes dominant.

@@

As seen from the RF-diagram in \figref{RF-O3a}, `xu' is no
longer used for \emph{O3}. Instead the word `pasi' comes to 
dominate. `pasi' has indeed the more specific 
meaning [\textsc{g} 0.375–0.5] which is only applicable to \emph{O3}, 
not to \emph{O5}. At the same time we see from the RF-diagram 
for \emph{O5} (\figref{RF-O5})
that the word `rimebi' initially dominates for designating \emph{O5}. 
`rimebi' has the more specific meaning [\textsc{g} 0.25–0.375] which 
is not applicable to \emph{O3}. 
The more general word `xu' is still useful in some 
contexts where the
refined distinction is not necessary (for example where 
either \emph{O3} or \emph{O5} is present but not both). So we would expect
that `xu' continues to exist. However this is not the case. `xu'
loses out completely and its role is taken over by `fepi'. 
\clearpage
Understanding why this is the case
tells us a great detail about the kind 
of hidden semiotic dynamics that takes place. `xu' loses its
strength because (1) it fails in games where its meaning
is \todo{inserted `is' here} not distinctive enough so that its score goes
down, and (2) because there 
are other meanings competing with `xu' which do not 
have strong alternatives and are therefore
less prone to failure. This is notably the case for
`fepi'. `fepi' carries the more general meaning of green
and does not have competitors. It therefore overtakes
`xu'.

>>>>

These competition diagrams are an important tool to try 
and make sense of the ontological and lexical evolution
taking place in evolving groups of agents as they are
playing their language games. Typically we pick
a dominating word, for example the word `va' in 
\figref{RF-diagram}, and try to understand
why it has become dominant. 
The FM-diagram for `va' (\figref{FM-diagram}) explains
part of the story. Three stable meanings for `va' have emerged
at  around 1000 games: 
[\textsc{r} 0.0–0.125], [\textsc{b} 0.3125–0.375], and [\textsc{vpos} 0.25–0.5]. 
They are all equally adequate for distinguishing the object
`va' designates, and there are no situations yet
that would have forced the disambiguation of `va'. 
In game 3000, the environment (which is continuously 
changing in this experiment) produces a scene in which
a category which was distinctive for the object designated by 
`va' is no longer distinctive. The lexicon adapts 
immediately. Around game 3000 the \textsc{vpos}-based meaning disappears, 
and the distinction based on \textsc{red} shoots up and becomes dominant.

@@

As seen from the RF-diagram in \figref{RF-O3a}, `xu' is no
longer used for \emph{O3}. Instead the word `pasi' comes to 
dominate. `pasi' has indeed the more specific 
meaning [\textsc{g} 0.375–0.5] which is only applicable to \emph{O3}, 
not to \emph{O5}. At the same time we see from the RF-diagram 
for \emph{O5} (\figref{RF-O5})
that the word `rimebi' initially dominates for designating \emph{O5}. 
`rimebi' has the more specific meaning [\textsc{g} 0.25–0.375] which 
is not applicable to \emph{O3}. 
The more general word `xu' is still useful in some 
contexts where the
refined distinction is not necessary (for example where 
either \emph{O3} or \emph{O5} is present but not both). So we would expect
that `xu' continues to exist. However this is not the case. `xu'
loses out completely and its role is taken over by `fepi'. 
\clearpage
Understanding why this is the case
tells us a great detail about the kind 
of hidden semiotic dynamics that takes place. `xu' loses its
strength because (1) it fails in games where its meaning
is not distinctive enough so that its score goes
down, and (2) because there 
are other meanings competing with `xu' which do not 
have strong alternatives and are therefore
less prone to failure. This is notably the case for
`fepi'. `fepi' carries the more general meaning of green
and does not have competitors. It therefore overtakes
`xu'.
